---
title: "R basics_20200906"
output: html_notebook
---

# 2.2 The very basics

Before we get started with the motivating dataset, we need to cover the very basics of R.

### 2.2.1 Objects

Suppose a high school student asks us for help solving several quadratic equations of the form $ax^2+bx+c = 0$. The quadratic formula gives us the solutions:

$$
\frac{-b - \sqrt{b^2 - 4ac}}{2a}\,\, \mbox{ and } \frac{-b + \sqrt{b^2 - 4ac}}{2a}
$$
which of course change depending on the values of $a$, $b$ and $c$. One advantage of programming languages is that we can define variables and write expressions with these variables, similar to how we do so in math, but obtain a numeric solution. We will write out general code for the quadratic equation below, but if we are asked to solve $x^2 + x -1 = 0$, then we define:

```{r}
a <- 1
b <- 1
c <- -1
```

which stores the values for later use. We use `<-` to assign values to the variables.We can also assign values using `=` instead of `<-`, but we recommend against using `=` to avoid confusion.

Copy and paste the code above into your console to define the three variables. Note that R does not print anything when we make this assignment. This means the objects were defined successfully. Had you made a mistake, you would have received an error message.

To see the value stored in a variable, we simply ask R to evaluate a and it shows the stored value:


```{r}
a
```
A more explicit way to ask R to show us the value stored in `a` is using `print` like this:


```{r}
print(a)
```
We use the term object to describe stuff that is stored in R. Variables are examples, but objects can also be more complicated entities such as functions, which are described later.


###2.2.2 The workspace

As we define objects in the console, we are actually changing the workspace. You can see all the variables saved in your workspace by typing:

```{r}
ls()
```
In RStudio, the Environment tab shows the values:

We should see `a`, `b`, and `c`. If you try to recover the value of a variable that is not in your workspace, you receive an error. For example, if you type `x` you will receive the following message: `Error: object 'x' not found`.

Now since these values are saved in variables, to obtain a solution to our equation, we use the quadratic formula:

```{r}
(-b + sqrt(b^2 - 4*a*c) ) / ( 2*a )
#> [1] 0.618
(-b - sqrt(b^2 - 4*a*c) ) / ( 2*a )
#> [1] -1.62
```
###2.2.3 Functions

Once you define variables, the data analysis process can usually be described as a series of functions applied to the data. R includes several predefined functions and most of the analysis pipelines we construct make extensive use of these.




